---
title: "ENV 797 - Time Series Analysis for Energy and Environment Applications | Spring 2025"
subtitle: "Assignment 6 - Due date 02/27/25"
author: "Janice Ye"
output: pdf_document
geometry: margin=2.54cm
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A06_Sp25.Rmd"). Then change "Student Name" on line 4 with your name.

Then you will start working through the assignment by **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai.

R packages needed for this assignment: "ggplot2", "forecast", "tseries" and "sarima". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.

```{r}
#Load/install required package here
library(ggplot2)
library(forecast)
library(tseries)

install.packages("sarima")
library(sarima)

```

This assignment has general questions about ARIMA Models.

## Q1

Describe the important characteristics of the sample autocorrelation function (ACF) plot and the partial sample autocorrelation function (PACF) plot for the following models:

* AR(2)

> Answer: ACF tails off gradually instead of cutting off at a fixed lag, while PACF cuts off after lag 2 with spikes at lag 1 and 2.

* MA(1)

> Answer: ACF has a spike at lag 1 then cuts off afterwards. PACF will tail off gradually.

## Q2

Recall that the non-seasonal ARIMA is described by three parameters ARIMA$(p,d,q)$ where $p$ is the order of the autoregressive component, $d$ is the number of times the series need to be differenced to obtain stationarity and $q$ is the order of the moving average component. If we don't need to difference the series, we don't need to specify the "I" part and we can use the short version, i.e., the ARMA$(p,q)$.

(a) Consider three models: ARMA(1,0), ARMA(0,1) and ARMA(1,1) with parameters $\phi=0.6$ and $\theta= 0.9$. The $\phi$ refers to the AR coefficient and the $\theta$ refers to the MA coefficient. Use the `arima.sim()` function in R to generate $n=100$ observations from each of these three models. Then, using `autoplot()` plot the generated series in three separate graphs.

```{r}
#parameter
n <- 100
phi <- 0.6
theta <- 0.9

#AR
ar1 <- arima.sim(n = n, model = list(ar = phi))

ar1_plot <- autoplot(ar1) +
  labs(title = "Simulated ARMA(1,0)") +
  theme(plot.title = element_text(hjust = 0.5))
ar1_plot

#MA
ma1 <- arima.sim(n = n, model = list(ma = theta))

ma1_plot <- autoplot(ma1) +
  labs(title = "Simulated ARMA(0,1)") +
  theme(plot.title = element_text(hjust = 0.5))
ma1_plot

#ARMA
arma11 <- arima.sim(n = n, model = list(ar = phi, ma = theta))

arma11_plot <- autoplot(arma11) +
  labs(title = "Simulated ARMA(1,1)") +
  theme(plot.title = element_text(hjust = 0.5))
arma11_plot

```

(b) Plot the sample ACF for each of these models in one window to facilitate comparison (Hint: use `cowplot::plot_grid()`).


```{r}
ar_acf <- ggAcf(ar1, lag.max = 30) +
  labs(title = "AR(1,0) ACF") +
  theme(plot.title = element_text(hjust = 0.5))
ar_acf

ma_acf <- ggAcf(ma1, lag.max = 30) +
  labs(title = "MA(0,1) ACF") +
  theme(plot.title = element_text(hjust = 0.5))
ma_acf

arma_acf <- ggAcf(arma11, lag.max = 30) +
  labs(title = "ARMA(1,1) ACF") +
  theme(plot.title = element_text(hjust = 0.5))
arma_acf

cowplot::plot_grid(ar_acf, ma_acf, arma_acf,
                   nrow = 1, align = "h")
```

(c) Plot the sample PACF for each of these models in one window to facilitate comparison.

```{r}
ar_pacf <- ggPacf(ar1, lag.max = 30) +
  labs(title = "AR(1,0) PACF") +
  theme(plot.title = element_text(hjust = 0.5))
ar_pacf

ma_pacf <- ggPacf(ma1, lag.max = 30) +
  labs(title = "MA(0,1) PACF") +
  theme(plot.title = element_text(hjust = 0.5))
ma_pacf

arma_pacf <- ggPacf(arma11, lag.max = 30) +
  labs(title = "ARMA(1,1) PACF") +
  theme(plot.title = element_text(hjust = 0.5))
arma_pacf

cowplot::plot_grid(ar_pacf, ma_pacf, arma_pacf,
                   nrow = 1, align = "h")

```

(d) Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be able identify them correctly? Explain your answer.

> Answer: In theory, AR(1)'s ACF tails off with PACF cuts off at lag 1. Based on my plot, The PACF shows a clear significant spike at lag 1, with the rest mostly small. With this signal, we can identify the model correctly. In theory, MA(1)'s ACF cuts off at lag 1 and PACF tails off. Based on my plot, I got a strong lag-1 ACF spike, which points to MA(1). But PACF's cutoff look less clean. With the ACF has a standout lag 1, we can identify the model correctly. In theory, ARMA(1,1)'s ACF and PACF tails off. Based on my plot, there are a few strong early lags, then a gradual decline. But since ARMA(1,1) can look like AR(1) or MA(1) in small samples because the AR and MA effects can partly offset or imitate each other at low lags, we cannot correctly identify this model. 

(e) Compare the PACF values R computed with the values you provided for the lag 1 correlation coefficient, i.e., does $\phi=0.6$ match what you see on PACF for ARMA(1,0), and ARMA(1,1)? Should they match?

> Answer: In theory, ARMA(1,0)'s PACF at lag 1 should equal to 0.6. But based on my plot, it’s not exactly 0.6 since it is an estimate based on n = 100, so it will deviate but in the neighborhood of 0.6. In theory, ARMA(1,1)'s PACF should not equal to 0.6 because it is based on both phi and theta. 

(f) Increase number of observations to $n=1000$ and repeat parts (b)-(e).

```{r}
#parameter
n1 <- 1000
phi1 <- 0.6
theta1 <- 0.9

#AR
ar1_new <- arima.sim(n = n1, model = list(ar = phi1))

ar1_new_plot <- autoplot(ar1_new) +
  labs(title = "Simulated ARMA(1,0)") +
  theme(plot.title = element_text(hjust = 0.5))
ar1_new_plot

#MA
ma1_new <- arima.sim(n = n1, model = list(ma = theta1))

ma1_new_plot <- autoplot(ma1_new) +
  labs(title = "Simulated ARMA(0,1)") +
  theme(plot.title = element_text(hjust = 0.5))
ma1_new_plot

#ARMA
arma11_new <- arima.sim(n = n1, model = list(ar = phi1, ma = theta1))

arma11_new_plot <- autoplot(arma11_new) +
  labs(title = "Simulated ARMA(1,1)") +
  theme(plot.title = element_text(hjust = 0.5))
arma11_new_plot

```

## Q3

Consider the ARIMA model $y_t=0.7*y_{t-1}-0.25*y_{t-12}+a_t-0.1*a_{t-1}$

(a) Identify the model using the notation ARIMA$(p,d,q)(P,D,Q)_ s$, i.e., identify the integers $p,d,q,P,D,Q,s$ (if possible) from the equation.

> Answer: ARIMA(1,0,1)(1,0,0)12. p = 1 from 0.7*y_{t-1} which is a nonseasonal AR(1); -0.25*y_{t-12} is a seasonal AR term at lag 12, so P = 1 and s = 12; -0.1*a_{t-1} is a nonseasonal MA(1) term, meaning q = 1. Since there is no seasonal MA term, Q = 0.

(b) Also from the equation what are the values of the parameters, i.e., model coefficients.

> Answer: phi = 0.7 which is the coefficient from y_{t-1}; Phi = -0.25 which is the coefficient from y_{t-12}; theta = -0.1 which is the coefficient from a_{t-1}.

## Q4

Simulate a seasonal ARIMA$(0, 1)\times(1, 0)_{12}$ model with $\phi =0 .8$ and $\theta = 0.5$ using the `sim_sarima()` function from package `sarima`. The $12$ after the bracket tells you that $s=12$, i.e., the seasonal lag is 12, suggesting monthly data whose behavior is repeated every 12 months. You can generate as many observations as you like. Note the Integrated part was omitted. It means the series do not need differencing, therefore $d=D=0$. Plot the generated
series using `autoplot()`. Does it look seasonal?

```{r}
seasonal_ARIMA <- sim_sarima(n = 120, model = list(ma = 0.5, sar = 0.8, nseasons = 12))

seasonal_ARIMA_ts <- ts(seasonal_ARIMA, frequency = 12)

autoplot(seasonal_ARIMA_ts) +
  ggtitle("Simulated ARIMA(0,0,1)x(1,0,0) [12] Process") +
  xlab("Year") +
  ylab("Value") +
  theme(plot.title = element_text(hjust = 0.5))

```

> Answer: Yes, it looks seasonal because Phi is very high at 0.8 as we see echos in data with cyclical peaks at the beginning of a year over 12-month period.

## Q5

Plot ACF and PACF of the simulated series in Q4. Comment if the plots are well representing the model you simulated, i.e., would you be able to identify the order of both non-seasonal and seasonal components from the plots? Explain.

```{r}
seasonal_arima_acf  <- ggAcf(seasonal_ARIMA_ts, lag.max = 60) +
  labs(title = "Sample ACF", x = "Lag", y = "ACF")

seasonal_arima_pacf <- ggPacf(seasonal_ARIMA_ts, lag.max = 60) +
  labs(title = "Sample PACF", x = "Lag", y = "PACF")

cowplot::plot_grid(seasonal_arima_acf, seasonal_arima_pacf, ncol = 1, align = "v")

```
> Answer: The ACF has large, significant spikes at lags 12 and its multiple, and they decrease gradually as lag increases. This is a signal of seasonality at 12 months. The PACF has one very large spike at lag 12, and then it’s mostly within the bounds at other seasonal multiples. This is what we expect from a seasonal AR(1). But we do not see the ACF cut off after lag 1 which makes q = 1 hard to identify. Overall, we could identify s = 12 and seasonal AR(1), but we could not identify non-seasonal MA(1).