---
title: "ENV 797 - Time Series Analysis for Energy and Environment Applications | Spring 2026"
subtitle: "Assignment 7 - Due date 03/06/26"
author: "Janice Ye"
output: pdf_document
geometry: margin=2.54cm
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A07_Sp26.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\

## Set up

```{r}
#Load/install required package here
library(forecast)
library(tseries)

```

## Importing and processing the data set

Consider the data from the file "Net_generation_United_States_all_sectors_monthly.csv". The data corresponds to the monthly net generation from January 2001 to December 2020 by source and is provided by the US Energy Information and Administration. **You will work with the natural gas column only**.


### Q1

Import the csv file and create a time series object for natural gas. Make you sure you specify the **start=** and **frequency=** arguments. Plot the time series over time, ACF and PACF.

```{r}
generation <- readr::read_csv(file = "./Data/Net_generation_United_States_all_sectors_monthly.csv",
                              skip = 4,
                              trim_ws = TRUE)
glimpse(generation)

#natural gas
ng <- generation %>% pull("natural gas thousand megawatthours")

ng_rev <- rev(ng)

ng_ts <- ts(ng_rev, start = c(2001,1), frequency = 12)

#Autoplot
autoplot(ng_ts) +
  labs(title = "Natural Gas Generation") +
  theme(plot.title = element_text(hjust = 0.5))

#ACF
og_acf <- ggAcf(ng_ts, lag.max = 40) +
  labs(title = "Natural Gas Generation ACF") +
  theme(plot.title = element_text(hjust = 0.5))
og_acf

#PACF
og_pacf <- ggPacf(ng_ts, lag.max = 40) +
  labs(title = "Natural Gas Generation ACF") +
  theme(plot.title = element_text(hjust = 0.5))
og_pacf

```

### Q2

Using the $decompose()$ and the $seasadj()$ functions create a series without the seasonal component, i.e., a deseasonalized natural gas series. Plot the deseasonalized series over time and corresponding ACF and PACF. Compare with the plots obtained in Q1.

```{r}
#decompose
ng_decompose <- decompose(ng_ts, type = "multiplicative")

#deseason
ng_deseason <- seasadj(ng_decompose)

#Autoplot
autoplot(ng_deseason, series = "Deseasoned") +
  labs(title = "Deseasoned Natural Gas Generation") +
  theme(plot.title = element_text(hjust = 0.5)) +
  autolayer(ng_ts, series = "Original")

#ACF
deseason_acf <- ggAcf(ng_deseason, series = "Deseasoned") +
  labs(title = "Deseasoned Natural Gas Generation PACF") +
  theme(plot.title = element_text(hjust = 0.5))
deseason_acf

cowplot::plot_grid(og_acf, deseason_acf,
                   nrow = 2,
                   align = "v")

#PACF
deseason_pacf <- ggPacf(ng_deseason, series = "Deseasoned") +
  labs(title = "Deseasoned Natural Gas Generation PACF") +
  theme(plot.title = element_text(hjust = 0.5))
deseason_pacf

cowplot::plot_grid(og_pacf, deseason_pacf,
                   nrow = 2,
                   align = "v")

```
> Answer: The original natural gas series in blue shows a clear upward trend from 2001 to 2020 with strong, repeating annual peaks and troughs, and the seasonal swings appear larger as the overall level rises. The deseasonalized series in red removes most of this recurring seasonal pattern, making the underlying growth trend much easier to see. Compared with the original, the deseasonalized line is smoother and highlights longer-term movements and irregular deviations without the seasonal spikes dominating the plot. In the original ACF, correlations remain strongly positive and decay slowly, with clear peaks at lags 12, 24, and 36 that indicate strong annual seasonality. After deseasonalizing, those seasonal peaks disappear, showing the seasonal component has been effectively removed. However, the deseasonalized ACF is still very high at short lags and declines gradually, suggesting the series remains highly persistent and likely still non-stationary due to trend. The original PACF has a very large spike at lag 1 and several additional significant spikes, indicating strong short-term dependence plus a seasonal structure. In the deseasonalized PACF, the seasonal related spikes largely vanish and most lags fall within the confidence bounds except at lag 1. This suggests that after removing seasonality, the remaining dependence is mainly short-run with weaker evidence of additional seasonal dynamics.

## Modeling the seasonally adjusted (i.e., the deseasonalized series)

### Q3

Run the ADF test and Mann Kendall test on the deseasonalized data from Q2. Report and explain the results.

```{r}
#ADF
adf.test(ng_deseason)

#MKT
mk.test(as.numeric(ng_deseason))

```

> Answer: For ADF test, the p-value is less than 0.05, which means that the null hypothesis is rejected indicating that the deseasonalized series is stationary and has no unit root. For MK test, the p-value is also less than 0.05, which means it has a significant positive monotonic trend.

### Q4

Using the plots from Q2 and test results from Q3 identify the ARIMA model parameters $p,d$ and $q$. Note that in this case because you removed the seasonal component prior to identifying the model you don't need to worry about seasonal component. Clearly state your criteria and any additional function in R you might use. DO NOT use the $auto.arima()$ function. You will be evaluated on ability to understand the ACF/PACF plots and interpret the test results.

```{r}
n_diff <- ndiffs(ng_deseason)
cat("Number of Differencing Needed", n_diff)

#check
deseason_n1 <- diff(ng_deseason, differences = 1, lag = 1)

#ACF
ggAcf(deseason_n1, lag.max = 40) +
  labs(title = "Differenced Deseasonalized ACF") +
  theme(plot.title = element_text(hjust = 0.5))

#PACF
ggPacf(deseason_n1, lag.max = 40) +
  labs(title = "Differenced Deseasonalized PACF") +
  theme(plot.title = element_text(hjust = 0.5))

```
> Answer: The number of differencing seems to be 1 because my plot in Q2 had a very slowly decaying ACF and Q3 typically shows ADF failing to reject a unit root. The function ndiffs() also proves the same. Based on the previous plot, the model seems to be ARIMA(1,1,0). In the differenced ACF, there is a strong negative spike at lag 1 and the remaining lags are mostly near the bounds. This is a classic signature of an MA(1) model. To account for the strong AR(1) relationship seen in the PACF, I set p = 1.

### Q5

Use `Arima()` from package "forecast" to fit an ARIMA model to your series considering the order estimated in Q4. You should allow constants in the model, i.e., `include.mean = TRUE` or `include.drift=TRUE`. **Print the coefficients** in your report. Hint: use the `cat()` or `print()` function to print.

```{r}
arima110 <- Arima(ng_deseason, order = c(1,1,0), include.drift = TRUE)

print(arima110)

```

> Answer: AR(1) is -0.1586 which is a weak, negative autoregressive term and suggests that we might be over-differencing. To fix the problem, I will now try ARIMA(1,0,0) to resolve the problem of over-differencing.

```{r}
arima100 <- Arima(ng_deseason, order = c(1,0,0), include.drift = TRUE)

print(arima100)
```

> Answer: This is a major improvement over the previous model. AR improves to a very strong, positive autoregressive coefficient. But just to double check on MA, I will run ARIMA (1,0,1). If MA(1)'s coefficient is small, I will keep with the ARIMA(1,0,0) model.

```{r}
arima101 <- Arima(ng_deseason, order = c(1,0,1), include.drift = TRUE)

print(arima101)
```

> Answer: This result is very clear that we should not inlcude MA(1) because the coefficient is too small. We will stick with ARIMA(1,0,0)

### Q6

Now plot the residuals of the ARIMA fit from Q5 along with residuals ACF on the same window. You may use the $checkresiduals()$ function to get the plots. Do the residual series look like a white noise series? Why?

```{r}
checkresiduals(arima100)

```
> Answer: It looks like a white-noise series because the residuals are centered around zero and show a relatively constant spread over the entire timeframe. here are no obvious cycles, trends, or large changing waves. However, the p-value is smaller than 0.05, rejecting null hypothesis that the serie is purely white noise. This means that there are remaining autocorrelation in the residuals.

## Modeling the original series (with seasonality)

### Q7

Repeat Q3-Q6 for the original series (the complete series that has the seasonal component). Note that when you model the seasonal series, you need to specify the seasonal part of the ARIMA model as well, i.e., $P$, $D$ and $Q$.

```{r}
#ADF
adf.test(ng_ts)

#MKT
mk.test(as.numeric(ng_ts))

```

> Answer: Both p-values from ADF and Mann-Kendall tests are smaller than 0.05. This means that we will reject null hypothesis that there is an unit-root and that there is no trend. In fact, the series is stationary and has an unit root. Since tau is 0.651, the serie follows an upward trend.

```{r}
n_diff_original <- ndiffs(ng_ts)
cat("Number of Differencing Needed", n_diff_original)

```

```{r}
original_n1 <- diff(ng_ts, differences = 1, lag = 12)

#ACF
ggAcf(original_n1, lag.max = 40) +
  labs(title = "Original ACF") +
  theme(plot.title = element_text(hjust = 0.5))

#PACF
ggPacf(original_n1, lag.max = 40) +
  labs(title = "Original PACF") +
  theme(plot.title = element_text(hjust = 0.5))
```

> Answer: Based on these ACF and PACF plots, possible ARIMA models are (1,0,0)(1,0,1). p = 1 because in PACF, there is a very strong, dominant spike at Lag 1, followed by a sharp cut-off which signals AR(1). d and q are 0 because there are no sharp cut-offs in ACF. P = 1 because in PACF, there are also significant spikes at Lag 12 and 13. This could suggest a Seasonal Autoregressive SAR(1) term, and there is a very prominent negative spike at Lag 12, and the spikes at Lags 24 and 36 are significantly smaller in ACF which suggests Q = 1.

```{r}
arima100101 <- Arima(ng_ts, order = c(1,0,0), seasonal = c(1,0,1), include.drift = TRUE)

print(arima100101)
```

> Answer: SAR(1) is 0.9919, which is extremely close to 1.0, which indicates that the seasonal component is non-stationary. It means that I should have used seasonal differencing ($D=1$) instead of trying to model the seasonality with an AR term. Now I would change seasonal from (1,0,1) to (0,1,1).

```{r}
arima100011 <- Arima(ng_ts, order = c(1,0,0), seasonal = c(0,1,1), include.drift = TRUE)

print(arima100011)
```

> Answer: This model improved a lot. AR(1) is 0.7416/0.0442 = 16.78, SMA(1) is -0.7026/0.0557 = -12.61, and drift is 358.7988/37.5875 = 9.55. These numbers are all significant and indicate a strong, positive relationship with the previous month's value. Now I need to double check with the residuals.

```{r}
checkresiduals(arima100011)

```

> Answer: P-value is 0.2777 and is greater than 0.05. I failed o reject the null hypothesis that the residuals are white noise.

### Q8

Compare the residual series for Q7 and Q6. Can you tell which ARIMA model is better representing the Natural Gas Series? Is that a fair comparison? Explain your response.

> Answer: The model from Q7 is better than Q6 because the test result from Q6 indicate that the model is failing to capture some systematic patterns whereas the result from Q7 fail to reject null hypothesis, indicating that the residuals are independent white noise. However, this is not a fair comparison because the first model is already deseasonalized while the second one is the original data. A fair comparison requires the models to be tested on the exact same target variable.

## Checking your model with the auto.arima()

**Please** do not change your answers for Q4 and Q7 after you ran the $auto.arima()$. It is **ok** if you didn't get all orders correctly. You will not loose points for not having the same order as the $auto.arima()$.

### Q9

Use the $auto.arima()$ command on the **deseasonalized series** to let R choose the model parameter for you. What's the order of the best ARIMA model? Does it match what you specified in Q4?

```{r}
deseason_test <- auto.arima(ng_deseason)

deseason_test

```

> Answer: The result shows that the best ARIMA model is ARIMA(1,1,1)(0,0,1). It does not match with mine.

### Q10

Use the $auto.arima()$ command on the **original series** to let R choose the model parameters for you. Does it match what you specified in Q7?

```{r}
original_test <- auto.arima(ng_ts)

original_test

```

> Answer: The result is ARIMA(1,0,0)(0,1,1) which perfectly match with my answer in Q7.